Research Experience

### 2010-03-15 doc/thesis/thesis.tex
work on particle filters+

@@

\subsection*{General Linear Model}
The most used form of FMRI analysis is still based on Statistical Parametric
Mapping, but is able to account for several different levels or types
of activation. By adding a General Linear Model to the analysis, the
output signal timeseries (what FMRI detects) is regressed over the weighted
sum of various confound timeseries. The equation for a general linear
model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of stimuli, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the inputs plus some error. The calculation
of $\beta$ then is merely a gradient descent search to minimize the
mean squared error.

@@

Of course, a square wave input is not going to result in a square wave
in the activation of brain regions. Thus, various methods are used to 
smooth $X(t)$ through time, and bandlimit the input. The best technique
is convolving the stimulus input with a hemodynamic response function,
which mimicks the basic shape of BOLD activation, including a delay
due to rise time and fall time. This hemodynamic signal is static however,
so every region of the brain still gets matched to the same signal,
although the weights of various stimulus or confounds (such as attention)
are allowed to vary.

@@

\subsection*{Whats wrong with these techniques}
There are a few problems with the techniques mentioned in the 
previous sections. First, they essentially ignore prior knowledge about
the system. The physiological models for the BOLD signal are quite good
and based in physics. Thus the parameters may be constrained by reality,
which crucially reduces model variance. Additionally, lacking real world
parameters means that the parameters have no clear physiological meaning,
so they are less useful in identifying pathologies.

Secondly, these methods are still merely t-tests, which notoriously
lack robustness to non-Gaussian noise. While different techniques 
exist for imposing Gaussianity, those techniques are incapable of
discriminating noise from signal. There is no way to know how much 
signal is removed
by various smoothing techniques, or even if entire regions have been
smoothed into oblivion. Instead of extensively filtering data to 
remove noise, the analysis method itself must be robust arbitrary
types of noise, which is why we propose here the use of particle filters.

\section*{Proposed Approach}
\subsection{Goal}
The ultimate goal of modeling neural activation is to learn something
new about how the brain works. The General Linear Model, and 
SPM's goal is to find a correlation between input, or what a 
person is thinking, and which regions are active. This is an 
important task, but it misses the potential of FMRI. FMRI's 
greatest weakeness as a tool for studying neural activity is that
it is connected to neurons through a long chain of events. But
what if, in addition to seeing which brain regions are active
for which stimuli, we could also study pathologies such as poor
blood flow, or derive physical parameters such as blood vessel
compliance? By using a realistic physical model as the basis for
discovering the relation between the outside world and FMRI 
signal, the path to detecting activation gives all those other
things for free, precisely \emph{because} FMRI is the result of such
a long chain of events. It is actually quite common to use models to derive
the values of physiological parameters. For instance, blood volume
is often measured by injecting a dye into the blood stream and then
assuming that the dye is removed at a rate proportional to the
concentration in the blood. By performing a simple regression it is
possible to estimate blood volume quite accurately. While for
this particular project I don't claim that the results will give 
physiologically plausible results, with some tuning of the algorithm
it would very possible.

>>>>

A large number of models have been proposed for the BOLD signal with varying
amounts of complexity. The simplest model is the so called "Balloon Model"
which first proposed the windkessel effect as a factor the BOLD response.
The model we will use is the model proposed by Buxton Et. Al. and later used
in Riera et. al. \cite{Riera}. The model has four state variables, $s$, $f$,
$v$, $q$, representing flow inducing signal, cerebral blood flow, cerebral
blood volume and deoxyhemaglobin to hemaglobin ratio, respectively. The 
state variables change over time, given by the state evolution equations:
\begin{equation}
\dot{s} = \epsilon u -  
\end{equation}

\begin{equation}
\dot{f} = s
\end{equation}

\begin{equation}
\dot{v} = 
\end{equation}

\begin{equation}
\dot{q} = 
\end{equation}

Additional papers have added such things as
....

@@

\subsection*{General Linear Model}
The most used form of FMRI analysis is still based on Statistical Parametric
Mapping, but is able to account for several different levels or types
of stimulus. By adding a General Linear Model to the analysis, the
output signal timeseries (what FMRI detects) is regressed over the weighted
sum of the various confound's timeseries. The equation for a general linear
model is then
\begin{equation}
Y(t) = X(t)\beta + \epsilon(t)
\end{equation}
where $Y(t)$ is the smoothed or detrended timeseries of measurements,
$X(t)$ is a row vector of stimuli, $\beta$ is a column vector of weights,
and $\epsilon$ is the error. Thus for every time, the measurement is
assumed to be a weighted sum of the inputs plus some error. The calculation
of $\beta$ then is merely a gradient descent search to minimize the
mean squared error.

@@

Of course, a square wave input is not going to result in a square wave
in the activation of brain regions. Thus, various methods are used to 
smooth $X(t)$ through time, and bandlimit the input. The best technique
is convolving the stimulus input with a hemodynamic response function,
which mimicks the basic shape of BOLD activation, including a delay
due to rise time and fall time. This hemodynamic signal is static however,
so every region of the brain gets the same design matrices (X(t)), 
although the weights of various stimulus or confounds are allowed to vary.

@@

\subsection*{Whats wrong with these techniques}
There are a few problems with the techniques mentioned in the 
previous sections. First, they essentially ignore prior knowledge about
the system. Although the most advanced form of the general linear model includes
a "Hemodynamic Response Function," that hemodynamic response function is
static across every region of the brain. It is well known that capillary beds
are not uniform and so blood perfusion cannot possible be static across the
brain. Thus, if extra information were available a-priori, that information
could not be incorporated without modifications to the General Linear Model.
Similarly heart rate could not be added either. It would obviously be advantageous
to have true physiological parameters as entry points for these various other
model parameters. The physiological models for the BOLD signal are quite good
and based on realistic physics. While the exact connection between a stimulus
and the flow inducing signal is not precisely known, model fits are actually
quite good \cite{ModelCompareison}. Regardless, being based on some real 
physiological 
parameter would allow for the establishment of reasonable priors and decrease
model variance without breaking a sweat. Of course, using real parameters 
has the additional
bonus of potentially providing information about physical pathologies. It
is quite possible that physical properties such as decreased compliance of
blood vessels could indicate a neurological condition that is not easily
seen in a T1 or T2 map. In essence, this could make FMRI a much more 
useful clinical tool than it is now. The other problem with linear models
is that they are a linear fit to a nonlinear signal. It is not uncommon
for data to be thrown out in FMRI studies because no significant activation
has been seen. However, if, for whatever reason, the BOLD response
was acting more nonlinear than in other patients it would be completely
possible for SPM to miss that activation.

<Image with two different $\alpha$s>

<image comparing the results of 10\% changes in various signals>

Secondly, these methods are still based on t-tests, which notoriously
lack robustness to non-Gaussian noise. While different techniques 
exist for imposing Gaussianity, those techniques are incapable of
discriminating noise from signal. There is no way to know how much 
signal is removed by various smoothing techniques, or even if entire 
regions have been smoothed into oblivion. Instead of extensively filtering data to 
remove noise, the analysis method itself must be robust a wide range of noise,
which is why we propose here the use of particle filters.

\section*{Proposed Approach}
\subsection*{Goal}
The ultimate goal of this project is to provide a new set of tools
for analyzing FMRI data. Whereas SPM techniques have been highly 
successful at finding macroscopic regions of activation, linear 
modeling can carry significant bias error due to lack of model
flexibility. While adding parameters can significantly increase
error due to model variance, this effect is mitigated by the fact
that we plan to use a model that is based on first principals. The
purpose of this paper is thus to evaluate the potential of using
a particle filter along with the BOLD model to derive physical 
parameters. In so doing, we hope to be able to show that neuronal
efficacy, $\epsilon$ is a sutiable variable for estimating voxel 
activation from a standard FMRI image. We also hope to show that 
estimated posterior distribution of the parameters, derived from
the particle filter, is able to provide an accurate measure of the
confidence interval.

\subsection*{Introduction to Particle Filters}
Particle filters, also known as Sequential Monte-Carlo methods (SMC),
are essentially an on-line version of the Markov-Chain Monte Carlo (MCMC)
algorithm. Similar to MCMC, Particle Filters assume the Markov Assumption
holds, in essence that 
\begin{equation}
P(X_t | X_{t-1}, X_{t-2}, ... X_{0}) = P(X_t | X_{t-1})$
\end{equation}
for a time evolving parameter, $X_t$. This is actually a very good
assumption in most biological systems, since chemicals driving physiological
parameters rarely have memory. Unlike MCMC algorithms, particle filters
only calculate the posterior distribution:
\begin{equation}
P(X_t | y_t, y_{t-1}, ... y_0)
\end{equation}
for $t$ measurements, $y_t, ... y_0$. This frees the algorith from having
to calculate, or even keep track of the values of $X$ at times other than
$t$, and significantly reducing computation time. This also easily lends
the particle filter to solving stochastic differential equations (SDE's).

...

\subsection*{My Specific Particle Filte}
Although particle filters are well defined, significant design decisions
are application dependent. The most important choice is the weighting function,
$w(y_t - y(\theta_t))$. Obviously the function should weight $0$ maximally,
the drop off rate is extremely important. A very thin gaussian probability 
distribution function has nice properties, but thin tails. As a result, large
outliers in the measurement vector could easily force all the particles to
have near 0 weights, thus forcing the particle filter to converge improperly. 
On the other end of the spectrum, a cauchy pdf may not
weight particles in the middle enough, preventing the particles to never 
converge. Of course, the scale factor, or variance, also plays a signifcant
role in the convergeance rate. The importance of the weighting function
cannot be overstated, as this is the primary factor in deciding the rate at
which the particle filter converges. As part of the simulations, I compare
a gaussian based weighting function with an exponential weightin function.
In both cases, I scaled the width of the weighting functions automatically based
on the signal variance.

Dealing with low frequency noise is also a crucial part of the tests. 
I tested two different methods of dealing with this noise. Various methods
are used to deal with this problem in other toolsets. The most common method
is to apply a high-pass filter to the measurement vector. Although this makes
sense in many applications, in this case it may not be adaptive enough for
a random walk type noise. On the other hand, it has been shown that a spline
de-trending method is able to better deal with the low frequency noise
typically present in FMRI. Thus, my first method
was to use a spline fit with knots every 80 or 90 measurements to trance
the basic shape of the drift. I then added an extra DC gain parameter to
the model to account for the fact that the de-trended signal tended to have 
a mean of 0, rather than some positive value.

My second method of dealing with low frequency drift was to counteract
it by weighting particles based on $(y_t-y_{t-1}) - (y(\theta_t)-y(\theta_{t_1}))$ 
rather than directly on $y_t - y(\theta_t)$. It is generally believed that the noise present in 
FMRI has approximately gaussian steps, which means that this method only
has to deal with gaussian white noise. Such a technique is usually not considered a
good idea because high frequency noise is almost exclusively assumed in
systems theory. In this case however; the noise is all in the lower frequency
range, so performing a differentiation like operation, which magnifies high
frequency signals, is exactly what we want.

... more specifics...

The General Linear Model, and 
SPM's goal is to find a correlation between input, or what a 
person is thinking, and which regions are active. This is an 
important task, but it misses the potential of FMRI. FMRI's 
greatest weakeness as a tool for studying neural activity is that
it is connected to neurons through a long chain of events. But
what if, in addition to seeing which brain regions are active
for which stimuli, we could also study pathologies such as poor
blood flow, or derive physical parameters such as blood vessel
compliance? By using a realistic physical model as the basis for
discovering the relation between the outside world and FMRI 
signal, the path to detecting activation gives all those other
things for free, precisely \emph{because} FMRI is the result of such
a long chain of events. It is actually quite common to use models to derive
the values of physiological parameters. For instance, blood volume
is often measured by injecting a dye into the blood stream and then
assuming that the dye is removed at a rate proportional to the
concentration in the blood. By performing a simple regression it is
possible to estimate blood volume quite accurately. While for
this particular project I don't claim that the results will give 
physiologically plausible results, with some tuning of the algorithm
it would very possible.
